ManyEcoEvo: an R package for working with the "Same Data, Different Analysts" Project in Ecology and Evolutionary Biology, and other many-analyst style data

### 2023-11-15 manuscript/manuscript/manuscript.Rmd
close #25 incorporate FF feedback

Typical scientific approaches proceed with a single team, and one analytic specification, or at most several specifications implemented.
Thus the effect of alternative defensible analytic choices that could have been implemented is never known, either because alternative specifications aren't explored, or worse, because analysts / authors implement them but do not publish alternative analysis specifications.
However, we know from simulations and case studies that researcher degrees of freedom can result in spurious findings [@simmons2011], and these researcher degrees of freedom are susceptible to publication bias, p-hacking and other questionable research practice.
Similarly, specification curve analysis of published findings have demonstrated that the published conclusion had empirical support from only a small sample of the overall plausible specifications (evidence of spurious findings, and possibly publication bias and p-hacking) [@Simonsohn2019].

Many-analyst studies are a format of crowd-sourced scientific analysis [@schweinsberg2021], where analysts simultaneously, but independently, research the same research question, using the same dataset [@Voracek2019; @silberzahn2018].
Many-analyst studies demonstrate how subjective, often arbitrary and usually defensible analysis decisions may influence research results [@silberzahn2018] due to either disagreement among researchers or due to analytic uncertainty about choices such as which observations to include and what data-analysis strategy is most appropriate [@Voracek2019].

@@

All raw data sets and metadata provided to independent teams for analysiss are publicly archived at <https://osf.io/qjzby> and <https://osf.io/hdv8m>, respectively.
Analyst survey responses and results are provided in the package GitHub repository (<https://github.com/egouldo/ManyEcoEvo/>), which is also archived on Zenodo (<https://doi.org/10.5281/zenodo.10046153>).
*ManyEcoEvo* contains tools for data-cleaning and processing tasks encountered when working with with many-analyst style data, functions for implementing and reproducing the analysis described inGould *et al.* [-@gould2023], establishes a structured workflow that facilitates re-analyses and replications of Gould *et al.* [-@gould2023] as well as implementation for new many-analyst studies, including a planned extension of the ManyEcoEvo project.

@@

The *ManyEcoEvo* package was created to implement the analysis for the 'Same Data, Many Analysts in Ecology and Evolutionary Biology' project [@gould2023] and to allow the 246 co-authors and other interested researchers to computationally reproduce our analysis.
Following [@Marwick2018], [@Vuorre2021] and [@Filazzola2022], we opted to develop this research compendium as an R package as we find it best encapsulates the principles of 'clean code' for effective communication of the science, and therefore facilitates future reproduction, replication and potential synthesis [R packages as research compendia].
The results of some many-analyst studies have been controversial, yielding re-analyses criticising the original implementation of some many-analyst studies.
For example, [@Auspurg2021] re-analysed part of the seminal 'Red-Card' many-analyst study examining the influence of whether referees were more likely to give red cards to soccer players with dark skin tone than those with light skin tone [@silberzahn2018].
By providing the code and data for Gould *et al.* [-@gould2023] the *ManyEcoEvo* package, we aim to facilitate computational reproductions by others in order to verify our results.
We also aim to facilitate analytic reproductions of this study, wherein *ManyEcoEvo* can be used to re-run the analysis in Gould *et al.* [-@gould2023] with slightly different methods, or subsets of data in order to verify and examine the robustness of our findings.

The last ten years has seen the growing emergence of many-analyst projects, [e.g. @silberzahn2018; @botvinik-nezer2020; @huntington-klein2021; @schweinsberg2021; @breznau2022; @hoogeveen2022many; @coretta2023], predominantly in the disciplines of psychology and social and behavioural sciences, however we expect more of these types of studies to be published into the near future across other disciplines.
Moreover, collaborative and distributed big team open science projects are becoming increasingly common in ecology and evolutionary biology (e.g. NutNet/DragNet <https://nutnet.org/index.php/>, ManyBirds <http://themanybirds.com>) as well as other disciplines [ManyBabies @ManyBabies , ManyPrimates @ManyPrimates] which pose computationally analogous data-manipulation and analysis problems to the many-analyst workflow addressed by the *ManyEcoEvo* approach. While *ManyEcoEvo* was designed for the ManyEcoEvo project, the structured approach it employs establishes a workflow that can be generalised and applied to new many-analyst style data, facilitating replication of Gould *et al.* [-@gould2023], in addition to computational reproduction [@Stoudt2021], as well as new analyses of many-analyst type studies.

>>>>

Many-analyst studies are a format of crowd-sourced scientific analysis [@schweinsberg2021], where analysts simultaneously, but independently, research the same research question, using the same dataset [@Voracek2019; @silberzahn2018].
Many-analyst studies demonstrate how subjective, often arbitrary and usually defensible analysis decisions may influence research results [@silberzahn2018] due to either disagreement among researchers or due to analytic uncertainty about choices such as which observations to include and what data-analysis strategy is most appropriate [@Voracek2019].
Managing the analysis pipeline of such data is complex and is better implemented with dedicated software.

Typical scientific approaches proceed with a single team, and one analytic specification, or at most several specifications implemented.
Thus the effect of alternative defensible analytic choices that could have been implemented is never known, either because alternative specifications aren't explored, or worse, because analysts / authors implement them but do not publish alternative analysis specifications.
However, we know from simulations and case studies that researcher degrees of freedom can result in spurious findings [@simmons2011], and these researcher degrees of freedom are susceptible to publication bias, p-hacking and other questionable research practice.
Similarly, specification curve analysis of published findings have demonstrated that the published conclusion had empirical support from only a small sample of the overall plausible specifications (evidence of spurious findings, and possibly publication bias and p-hacking) [@Simonsohn2019].
These types of studies are growing in popularity, as we explain below.

@@

All raw data sets and metadata provided to independent teams for analysis are publicly archived at <https://osf.io/qjzby> and <https://osf.io/hdv8m>, respectively.
Analyst survey responses and results are provided in the package GitHub repository (<https://github.com/egouldo/ManyEcoEvo/>), which is also archived on Zenodo (<https://doi.org/10.5281/zenodo.10046153>).
*ManyEcoEvo* contains tools for data-cleaning and processing tasks encountered when working with with many-analyst style data, functions for implementing and reproducing the analysis described inGould *et al.* [-@gould2023], establishes a structured workflow that facilitates re-analyses and replications of Gould *et al.* [-@gould2023] as well as implementation for new many-analyst studies, including a planned extension of the ManyEcoEvo project.

@@

The last ten years has seen the growing emergence of many-analyst projects, [e.g. @silberzahn2018; @botvinik-nezer2020; @huntington-klein2021; @schweinsberg2021; @breznau2022; @hoogeveen2022many; @coretta2023], predominantly in the disciplines of psychology and social and behavioural sciences, however we expect more of these types of studies to be published into the near future across other disciplines.
Moreover, collaborative and distributed big team open science projects are becoming increasingly common in ecology and evolutionary biology (e.g. NutNet/DragNet [nutnet.org/index.php/](https://nutnet.org/index.php/){.uri}, ManyBirds [themanybirds.com](http://themanybirds.com){.uri}) as well as other disciplines [ManyBabies @ManyBabies , ManyPrimates @ManyPrimates] which pose computationally analogous data-manipulation and analysis problems to the many-analyst workflow addressed by the *ManyEcoEvo* approach.

The *ManyEcoEvo* package was created to implement the analysis for the 'Same Data, Many Analysts in Ecology and Evolutionary Biology' project [@gould2023] and to allow the 246 co-authors and other interested researchers to computationally reproduce our analysis.
Following [@Marwick2018], [@Vuorre2021] and [@Filazzola2022], we opted to develop this research compendium as an R package as we find it best encapsulates the principles of 'clean code' for effective communication of the science, and therefore facilitates future reproduction, replication and potential synthesis [R packages as research compendia].
*ManyEcoEvo* was designed for the ManyEcoEvo project, the structured approach it employs establishes a workflow that can be generalised and applied to new many-analyst style data, facilitating replication of Gould *et al.* [-@gould2023], in addition to computational reproduction [@Stoudt2021], as well as new analyses of many-analyst type studies.

The results of some many-analyst studies have been controversial, yielding re-analyses criticising the original implementation of some many-analyst studies.
For example, [@Auspurg2021] re-analysed part of the seminal 'Red-Card' many-analyst study examining the influence of whether referees were more likely to give red cards to soccer players with dark skin tone than those with light skin tone [@silberzahn2018].
By providing the code and data for Gould *et al.* [-@gould2023] the *ManyEcoEvo* package, we aim to facilitate computational reproductions by others in order to verify our results.
We also aim to facilitate analytic reproductions of this study, wherein *ManyEcoEvo* can be used to re-run the analysis in Gould *et al.* [-@gould2023] with slightly different methods, or subsets of data in order to verify and examine the robustness of our findings.
